A Formal Model of Checked C

### 2022-07-16 sandp23/formal.tex
update.

There are two notions of type in \lang.  Types $\tau$ classify
word-sized values including integers and pointers, while types
$\omega$ classify multi-word values such as arrays, null-terminated
arrays, functions, and single-word-size values.
%
Pointer types ($\tptr{\omega}{m}$) include a mode annotation ($m$)
which is either checked (\cmode) or unchecked (\umode) and a type
($\omega$) denoting valid values that can be pointed to. Array types include both the type of
elements ($\tau$) and a bound ($\bvar$) comprised of an upper and
lower bound on the size of the array ($(b_l,b_h)$). Bounds $b$ are
limited to integer literals $n$ and expressions $x + n$.
Whether an array pointer is null terminated or not is determined by annotation
$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
otherwise (we elide the $\cdot$ when writing the type).
\systemname function types ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$)
reflect its dependent function declrations,
where $\overline{x}$ represents
a list of \tint type variables in a dependent function header,
which bind all bounds appearing in $\overline{\tau}$ and $\tau$.
We have a well-formed requirement for a function type,
such that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
Here is the
corresponding \checkedc syntax for these types:
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{array_tptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{nt_array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\\[0.2em]
$\code{tptr<(int)(nt_array_tptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{nt_array_tptr<}$\tau$\code{>)>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
As a convention we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

@@

\lang aims to be simple enough to work with, but powerful enough to
encode realistic \checkedc idioms. For example, mutable local
variables can be encoded as immutable locals that point to the heap;
the use of \code{&} can be simulated with \code{malloc};
and loops can be encoded as recursive function calls. \code{struct}s are
not in Fig.~\ref{fig:checkc-syn} for space reasons, but they are
actually in our model, and developed in
\iftr
Appendix~\ref{appx:struct}.
\else
the supplemental report~\cite{checkedc-tech-report}.
\fi
C-style \code{union}s have no safe typing
in \checkedc, so we omit them.
Additional syntactic descrition is listed in the \checkedc formalism paper \cite{li22checkedc}.

\ignore{
 By default, functions are assumed to
be within checked regions; placing the body in an \code{unchecked}
expression relaxes this, and within that, checked regions can be
nested via function calls. Bounds are restricted slightly:
rather than allowing arbitrary sub-expressions, bounds must be either
integer literals or variables plus an integer offset, which accounts
for most uses of \code{bounds} in \checkedc programs. \lang bounds are
defined as relative offsets, not absolute ones, as in the second part
of Fig.~\ref{fig:checkedc-example}. We see no technical problem to
modeling absolute bounds, but it would be a pervasive change so we
have not done so.

  We have mechanized two models of \lang, one in Coq
  and one in PLT Redex~\cite{pltredex}, which is a semantic
  engineering framework implemented in Racket. Redex provides direct
  support for specifying the operational semantics and typing with logical
  rules, but then automatically makes them executable and subject to
  randomized testing, which is very useful during development.
  The model we present in the paper faithfully
  represents both mechanizations, but there are some differences for
  presentation purposes. For example, the paper and the Coq model use an explicit
  stack, whereas the Redex model uses \code{let} bindings to simulate
  one (simplifying term generation for randomized
  testing).
\iftr
  Appendix~\ref{app:model-diffs} 
\else
  The supplemental report~\cite{checkedc-tech-report}
\fi
  outlines the differences
  between the two models and the paper formalism.
}
% \mwh{NOTE: COULD WORK THE FOLLOWING INTO THE ABOVE DESCRIPTION OF
%   POINTERS: Array and NT-array types have two relative bounds, whose structures
% can be either an integer or a variable plus an integer. For example,
% if we have the expression
% $x\texttt{=}\emalloc{\tarrayb{(0,10)}{\tint}}$, $x$ then has the type
% $\tptr{\tarrayb{10}{\tint}}{\cmode}$, which represents an array
% pointer of size $10$. The bounds ($(b_l,b_h)$ in
% $\evalue{p}{\tptr{\tarrayb{(b_l,b_h)}{\tau}}{m}}$) indicate relative
% offsets from this pointer of the accessible memory, i.e., $p+b_l$ and
% $p+b_h$. As such, a pointer can only be directly dereferenced if 0 is
% included within the annotated range.  If following the $\emalloctext$
% operation, we execute $*(x-1)$ and $*(x\plus 10)$. The two expressions
% are not valid, because the type of the two expressions $(x-1)$ and
% $(x\plus 10)$ are $\tarrayptr{1}{11}{\tint}{\cmode}$ and
% $\tarrayptr{-10}{0}{\tint}{\cmode}$ and $0$ in these two cases are not
% in the ranges: $[1,11)$ and $[-10,0)$. Thus, in order for a $\cmode$
% pointer with type $\tarrayptr{b_l}{b_h}{\tint}{\cmode}$ to be
% accessible in \checkedc, $0$ must be in the range of $[b_l,b_h)$.}

@@

    \inferrule{ m=\mode(E) \\
      e=E[\eif{\estar{x}}{e_1}{e_2}] \\
      (\varphi,\heap,\eif{\estar{x}}{e_1}{e_2})
              \longrightarrow (\varphi',\heap',e')}
    {(\varphi,\heap,e)\longrightarrow_{m} (\varphi',\heap',E[e'])}
    \;\;\;{[\emph{prefer}]}

@@

While heap bindings can change, stack bindings are immutable---once
variable $x$ is bound to $\evalue{n}{\tau}$ in $\varphi$, that binding will not
be updated; we can model mutable stack variables as pointers into the
mutable heap.
As mentioned, value $\evalue{0}{\tau}$
represents a $\enull$ pointer when $\tau$ is a pointer type;
correspondingly, $\heap(0)$ should always be undefined.
The relation steps to a \emph{result} $r$,
which is either an expression or a $\enull$ or $\ebounds$ failure,
representing a null-pointer dereference or out-of-bounds access,
respectively. Such failures are a \emph{good} outcome; stuck states
(non-value expressions that cannot transition to a result $r$)
characterize undefined behavior.
%
The context mode $m$ indicates whether the
stepped redex within $e$ was in a checked ($\cmode$) or
unchecked ($\umode$) region.

The rules for the main operational semantics
judgment---\emph{evaluation}---are given at the bottom of
Fig.~\ref{fig:c-context}.
The first rule takes an expression $e$, decomposes
it into an \emph{evaluation context} $E$ and a sub-expression $e'$
(such that replacing the hole $\Box$ in $E$ with $e'$ would yield
$e$), and then evaluates $e'$ according to the \emph{computation}
  relation $(\varphi,\heap,e') \longrightarrow (\varphi,\heap,e'')$,
whose rules are given in Fig.~\ref{fig:semantics}, discussed
shortly. 
The second rule handles conditionals $\eif{\estar{x}}{e_2}{e_3}$ in
redex position specially, delegating directly to the \textsc{S-IfNTT} computation
rule, which supports bounds widening; we discuss this rule
shortly. When the second and first rules could both apply, we always prefer the second.
The third rule describes the exception handling 
for possible crashing behaviors in unchecked region.
A program in $\umode$ mode can non-deterministically crash
and the \systemname sandbox mechanism recorvers
the program to a safe point ($\evalue{0}{\tau}$)
and continues with the existing program state.

 \begin{mathpar}
   \inferrule
       {}
       {m;\Theta;\heap;\sigma \vdash n : \tint}

   \inferrule
       {}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{\omega}{\umode}}

   \inferrule
       {}
       {m;\Theta;\heap;\sigma \vdash 0 : \tptr{\omega}{\xi}}

   \inferrule
       {(\evalue{n}{\tptr{\omega}{\xi}})\in \sigma}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{\omega}{\xi}}

   \inferrule
       {\tptr{\omega'}{\xi'} \sqsubseteq_{\Theta} \tptr{\omega}{\xi} 
            \\ m;\Theta;\heap;\sigma \vdash n : \tptr{\omega'}{\xi'}}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{\omega}{\xi}}

   \inferrule
       { \xi \le m 
     \\\Xi(m,n)=\tau\;(\evalue{\overline{x'}}{\overline{\tau}})\;(\xi,e)
       \\  \overline{x} = \{x|(x:\tint) \in (\overline{x'}:\overline{\tau}) \}}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{(\tfun{\overline{x}}{\overline{\tau}}{\tau})}{\xi}}

   \inferrule
       {\neg\funptr(\omega)\\ \xi \le m\\
        \forall i \in [0,\size(\omega)) \;.\;
            m;\Theta;\heap;(\sigma \cup \{(n:\tptr{\omega}{\xi})) \}\vdash \heap(m,n+i)}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{\omega}{\xi}}
 \end{mathpar}
 }
{\footnotesize
\[
\begin{array}{l} 
\funptr(\tfun{\overline{x}}{\overline{\tau}}{\tau}) = \texttt{true}
\qquad
\funptr(\omega) = \texttt{false}\;\;{[\emph{owise}]}
\end{array}
\\
\]
}
 \caption{Verification/Type Rules for Constants}
 \label{fig:const-type}
 \end{figure}
\end{DIFnomarkup}

%\footnote{This approach is that of the PLT Redex model of \lang; the Coq
%development uses a slightly simpler syntax to achieve the same
%effect.}
% \review{the special case raises questions, e.g. why is this syntax-driven and
%   not type-driven? }
% \liyi{This describes the semantic transition rules. We are using context evaluation framework to define the transition rules as the $E$ definition in Fig.3. like $\frac{x \Rightarrow y}{x+z \Rightarrow y + z}$, I don't know how type-driven can help us define translation rules.  }
% \mwh{Don't follow the above. I don't see this ``context transition
%   rule'' anywhere, and I'm not sure how it would fire, if we had it.}
% \liyi{The comment seems to confuse the meaning of the text about the if-then-else rules. Making the rule specific will help. }
The $\mode$ function
determines the context mode when evaluating $e'$ based on the context $E$.
For any program execution, the top context mode always starts with $\cmode$ ($\mode(E)=\mode'(E,\cmode)$).
The result context mode is depeneded on where the $\Box$ locates.
If it occurs within $(\eunchecked{\overline{x}}{E'})$ inside $E$ without
a surrounding $\echeckedtext$ block, then the mode is
$\umode$; if the exact opposite happens (within $(\echecked{\overline{x}}{E'})$ without surrounding $\euncheckedtext$ blocks),
the mode is $\cmode$; the other cases ($\mode'(\alpha({E}),m)=\mode'(E,m)$) are recursively defined by the above base cases ($\alpha$ represents any syntactic categories in \Cref{fig:c-context} other than $\Box$, $\euncheckedtext$, and $\echeckedtext$).
Evaluation contexts $E$ define a standard left-to-right evaluation order. (We explain the
$\ret{x}{\mu}{e}$ syntax shortly.)
% For a term $e$, $C$ is a context if any
% only if it contains a $\Box$ hole term in one of the subterm position
% in $e$. $\Box$ indicates the break point where we split $e$ into $C$
% and a subterm $e'$ such that $C[e'] = e$.  For a function call
% $\ecall{f}{\overline{e}}$, a valid context ($\overline{C}$) is defined
% as any one of the elements in the list $\overline{e}$ being a
% context. We show a rule of the upper level \checkedc semantics. The
% rule splits a term $e$ into a context $C$ and redex $e'$.  If subterm
% $e'$ is transitioned to $e_a$, $C[e']$ is transitioned to
% $C[e_a]$.

@@

        \inferrule[S-DefT]{\heap(\umode,n)=\evalue{n_a}{\tau_a}
         \\  \umode;\emptyset;\heap ; \emptyset \vdash\evalue{n_a}{\tau} }
    {(\varphi,\heap,\estar{\evalue{n}{\tptr{\tau}{\tmode}}}) \longrightarrow (\varphi,\heap,\evalue{n_a}{\tau})}

    \inferrule[S-AssignArrT]{\heap(\umode,n)=\evalue{n_a}{\tau_a}\\ 0 \in [n_l,n_h)
               \\  \umode;\emptyset;\heap ; \emptyset \vdash\evalue{n_1}{\tau} }
      {(\varphi,\heap,\eassign{\evalue{n}{\tallarrayptr{n_l}{n_h}{\tau}{\tmode}}}{\evalue{n_1}{\tau_1}}) \longrightarrow (\varphi,(\heap,\umode)[n \mapsto \evalue{n_1}{\tau_a}],\evalue{n_1}{\tau})}

@@

     \inferrule[S-IfNTTC]{\varphi(x)=\evalue{n}{\tntarrayptr{n_l}{0}{\tau}{\cmode}} \\ \heap(\cmode,n)\neq 0}
    {(\varphi,\heap,\eif{*x}{e_1}{e_2}) \longrightarrow (\varphi[x\mapsto \evalue{n}{\tntarrayptr{n_l}{1}{\tau}{\cmode}}],\heap,e_1)}

@@

        \inferrule[S-Let]{}{(\varphi,\heap,\elet{x}{\evalue{n}{\tau}}{e}) \longrightarrow (\varphi[x\mapsto \evalue{n}{\tau}],\heap,\ret{x}{\varphi(x)}{e})}

@@

    \inferrule[S-RetEnd]{}{(\varphi,\heap,\ret{x}{\mu}{\evalue{n}{\tau}}) \longrightarrow (\varphi,\heap,\evalue{n}{\tau})}

    \inferrule[S-Unchecked]{}{(\varphi,\heap,\eunchecked{\overline{x}}{\evalue{n}{\tau}}) \longrightarrow (\varphi,\heap,\evalue{n}{\tau})}

@@

    \inferrule[S-Checked]{}{(\varphi,\heap,\echecked{\overline{x}}{\evalue{n}{\tau}}) \longrightarrow (\varphi,\heap,\evalue{n}{\tau})}

    \inferrule[S-FunT]{ \Xi(\umode,n) = \tau\;(\evalue{\overline{x}}{\overline{\tau}})\;(\tmode,e)
                  \\ \umode;\emptyset;\heap ; \emptyset \vdash\evalue{n}{\tptr{\tau}{\tmode}}}
        {(\varphi,\heap,\ecall{\evalue{n}{(\tptr{\tau}{\tmode})}}{{\evalue{\overline{n_a}}{\overline{\tau_a}}}}) \longrightarrow
   (\varphi,\heap, \mathtt{let}\;\overline{x}={\evalue{\overline{n}}{(\overline{\tau}[\overline{n} / \overline{x}])}}\;\mathtt{in}\;\ecast{\tau[\overline{n} / \overline{x}]}{e})}

@@

@@

\myparagraph{Checked and Tainted Pointer Operations}
%
The rules for pointer related operations---\textsc{S-DefC},
\textsc{S-DefT}, \textsc{S-AssignArrC}, \textsc{S-AssignArrT},
\textsc{S-DefNull}, \textsc{S-IfNTTC} and \textsc{S-Cast}.
The first five are deference and assignment operations---illustrate how the semantics checks bounds.
Rule \textsc{S-DefNull} transitions attempted null-pointer
dereferences to $\enull$, whereas \textsc{S-DefC} dereferences a \cmode-mode
non-null (single) pointer.
  When $\enull$ is returned by the
computation relation, the evaluation relation halts the entire
evaluation with $\enull$ (using a rule not shown in Fig.~\ref{fig:c-context}); it
does likewise when $\ebounds$ is returned (see \Cref{sec:rem-semantics}).

@@

\textsc{S-AssignArrC} assigns to an array as long as 0 (the point of
dereference) is within the bounds designated by the pointer's annotation
and strictly less than the upper bound. For the assignment rule,
arrays are treated uniformly whether they are null-terminated or not
($\kappa$ can be $\cdot$ or $nt$)---the semantics does not search past
the current position for a null terminator. The program
can widen the bounds as needed, if they currently precede the null
terminator: \textsc{S-DefNTArray}, which dereferences an NT array
pointer, allows an upper bound of $0$, since the program may read, but
not write, the null terminator. A separate rule (not shown) handles
normal arrays.

\myparagraph{Casts}
%
Static casts of a literal $n\!:\!\tau'$ to a type $\tau$ are handled
by \textsc{S-Cast}. In a type-correct program, such casts are
confirmed safe by the type system. To evaluate a cast, the rule
updates the type annotation on $n$. Before doing so, it must
``evaluate'' any variables that occur in $\tau$ according to their
bindings in $\varphi$. For example, if $\tau$ was
$\tarrayptr{0}{x+3}{\tint}{\cmode}$, then $\varphi(\tau)$ would
produce $\tarrayptr{0}{5}{\tint}{\cmode}$ if $\varphi(x) = 2$.

Dynamic casts are accounted for by \textsc{S-DynCast} and
\textsc{S-DynCastBound}. In a type-correct program, such casts are
assumed correct by the type system, and later confirmed by the
semantics. As such, a dynamic cast will cause a $\ebounds$ failure if
the cast-to type is incompatible with the type of the target pointer,
as per the $n_l' > n_l \vee n_h > n_h'$ condition in
\textsc{S-DynCastBound}. An example use of dynamic casts is given on
line~7 in Fig.~\ref{fig:strcat-ex}. The values of \code{x} and
\code{n} might not be known statically, so the type system cannot
confirm that \code{x <= n}; the dynamic cast assumes this inequality
holds, but then checks it at run-time.

\myparagraph{Binding and Function Calls}
%
The semantics handles variable scopes using the special $\erettext$
form. \textsc{S-Let} evaluates to a configuration whose stack
is $\varphi$ extended with a binding for $x$, and whose expression is
$\ret{x}{\varphi(x)}{e})$ which remembers $x$ was previously bound to
$\varphi(x)$; if it had no previous binding, $\varphi(x) =
\bot$. Evaluation proceeds on $e$ until it becomes a literal
$n\!:\!\tau$, in which case \textsc{S-Ret} restores the saved
binding (or $\bot$) in the new stack, and evaluates to
$n\!:\!\tau$.

>>>>

There are two notions of type in \lang.  Types $\tau$ classify
word-sized values including integers and pointers, while types
$\omega$ classify multi-word values such as arrays, null-terminated
arrays, functions, and single-word-size values.
%
Pointer types ($\tptr{\omega}{m}$) include a mode annotation ($m$)
which is either checked (\cmode) or unchecked (\umode) and a type
($\omega$) denoting valid values that can be pointed to. Array types include both the type of
elements ($\tau$) and a bound ($\bvar$) comprised of an upper and
lower bound on the size of the array ($(b_l,b_h)$). Bounds $b$ are
limited to integer literals $n$ and expressions $x + n$.
Whether an array pointer is null terminated or not is determined by annotation
$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
otherwise (we elide the $\cdot$ when writing the type).
\systemname function types ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$)
reflect its dependent function declrations,
where $\overline{x}$ represents
a list of \tint type variables in a dependent function header,
which bind all bounds appearing in $\overline{\tau}$ and $\tau$.
We have a well-formed requirement for a function type,
such that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
Here is the
corresponding \systemname syntax for these types:
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{array_tptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{nt_array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\\[0.2em]
$\code{tptr<(int)(nt_array_tptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{nt_array_tptr<}$\tau$\code{>)>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
As a convention we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

@@

\lang aims to be simple enough to work with, but powerful enough to
encode realistic \systemname idioms. For example, mutable local
variables can be encoded as immutable locals that point to the heap;
the use of \code{&} can be simulated with \code{malloc};
and loops can be encoded as recursive function calls. \code{struct}s are
not in Fig.~\ref{fig:checkc-syn} for space reasons, but they are
actually in our model, and developed in
\iftr
Appendix~\ref{appx:struct}.
\else
the supplemental report~\cite{checkedc-tech-report}.
\fi
C-style \code{union}s have no safe typing
in \checkedc, so we omit them.
Additional syntactic descrition is listed in the \checkedc formalism paper \cite{li22checkedc}.

% \mwh{NOTE: COULD WORK THE FOLLOWING INTO THE ABOVE DESCRIPTION OF
%   POINTERS: Array and NT-array types have two relative bounds, whose structures
% can be either an integer or a variable plus an integer. For example,
% if we have the expression
% $x\texttt{=}\emalloc{\tarrayb{(0,10)}{\tint}}$, $x$ then has the type
% $\tptr{\tarrayb{10}{\tint}}{\cmode}$, which represents an array
% pointer of size $10$. The bounds ($(b_l,b_h)$ in
% $\evalue{p}{\tptr{\tarrayb{(b_l,b_h)}{\tau}}{m}}$) indicate relative
% offsets from this pointer of the accessible memory, i.e., $p+b_l$ and
% $p+b_h$. As such, a pointer can only be directly dereferenced if 0 is
% included within the annotated range.  If following the $\emalloctext$
% operation, we execute $*(x-1)$ and $*(x\plus 10)$. The two expressions
% are not valid, because the type of the two expressions $(x-1)$ and
% $(x\plus 10)$ are $\tarrayptr{1}{11}{\tint}{\cmode}$ and
% $\tarrayptr{-10}{0}{\tint}{\cmode}$ and $0$ in these two cases are not
% in the ranges: $[1,11)$ and $[-10,0)$. Thus, in order for a $\cmode$
% pointer with type $\tarrayptr{b_l}{b_h}{\tint}{\cmode}$ to be
% accessible in \checkedc, $0$ must be in the range of $[b_l,b_h)$.}

@@

@@

While heap bindings can change, stack bindings are immutable---once
variable $x$ is bound to $\evalue{n}{\tau}$ in $\varphi$, that binding will not
be updated; we can model mutable stack variables as pointers into the
mutable heap.
As mentioned, value $\evalue{0}{\tau}$
represents a $\enull$ pointer when $\tau$ is a pointer type;
correspondingly, $\heap(m,0)$ should always be undefined.
The relation steps to a \emph{result} $r$,
which is either an expression or a $\enull$ or $\ebounds$ failure,
representing a null-pointer dereference or out-of-bounds access,
respectively. Such failures are a \emph{good} outcome; stuck states
(non-value expressions that cannot transition to a result $r$)
characterize undefined behavior.
%
The context mode $m$ indicates whether the
stepped redex within $e$ was in a checked ($\cmode$) or
unchecked ($\umode$) region.

The rules for the main operational semantics
judgment---\emph{evaluation}---are given at the bottom of
Fig.~\ref{fig:c-context}.
The first rule takes an expression $e$, decomposes
it into an \emph{evaluation context} $E$ and a sub-expression $e'$
(such that replacing the hole $\Box$ in $E$ with $e'$ would yield
$e$), and then evaluates $e'$ according to the \emph{computation}
  relation $(\varphi,\heap,e') \longrightarrow (\varphi,\heap,e'')$,
whose rules are given in Fig.~\ref{fig:semantics}, discussed
shortly. 
The second rule describes the exception handling 
for possible crashing behaviors in unchecked region.
A program in $\umode$ mode can non-deterministically crash
and the \systemname sandbox mechanism recorvers
the program to a safe point ($\evalue{0}{\tau}$)
and continues with the existing program state.
There are other rules for describing the halts of evaluation to $\enull$ and $\ebounds$ states in \Cref{app:main}.

%\footnote{This approach is that of the PLT Redex model of \lang; the Coq
%development uses a slightly simpler syntax to achieve the same
%effect.}
% \review{the special case raises questions, e.g. why is this syntax-driven and
%   not type-driven? }
% \liyi{This describes the semantic transition rules. We are using context evaluation framework to define the transition rules as the $E$ definition in Fig.3. like $\frac{x \Rightarrow y}{x+z \Rightarrow y + z}$, I don't know how type-driven can help us define translation rules.  }
% \mwh{Don't follow the above. I don't see this ``context transition
%   rule'' anywhere, and I'm not sure how it would fire, if we had it.}
% \liyi{The comment seems to confuse the meaning of the text about the if-then-else rules. Making the rule specific will help. }
The $\mode$ function
determines the context mode when evaluating $e'$ based on the context $E$.
For any program execution, the top context mode always starts with $\cmode$ ($\mode(E)=\mode'(E,\cmode)$).
The result context mode is depeneded on where $\Box$ locates.
If it occurs within $(\eunchecked{\overline{x}}{E'})$ inside $E$ without
a surrounding $\echeckedtext$ block, then the mode is
$\umode$; if the exact opposite happens (within $(\echecked{\overline{x}}{E'})$ without surrounding $\euncheckedtext$ blocks),
the mode is $\cmode$; the other cases ($\mode'(\alpha({E}),m)=\mode'(E,m)$) are recursively defined by the above base cases ($\alpha$ represents any syntactic categories in \Cref{fig:c-context} other than $\Box$, $\euncheckedtext$, and $\echeckedtext$).
Evaluation contexts $E$ define a standard left-to-right evaluation order. (We explain the
$\ret{x}{\mu}{e}$ syntax shortly.)
% For a term $e$, $C$ is a context if any
% only if it contains a $\Box$ hole term in one of the subterm position
% in $e$. $\Box$ indicates the break point where we split $e$ into $C$
% and a subterm $e'$ such that $C[e'] = e$.  For a function call
% $\ecall{f}{\overline{e}}$, a valid context ($\overline{C}$) is defined
% as any one of the elements in the list $\overline{e}$ being a
% context. We show a rule of the upper level \checkedc semantics. The
% rule splits a term $e$ into a context $C$ and redex $e'$.  If subterm
% $e'$ is transitioned to $e_a$, $C[e']$ is transitioned to
% $C[e_a]$.

@@

        \inferrule[S-DefT]{\heap(\umode,n)=\evalue{n_a}{\tau_a}
         \\  \emptyset;\heap ; \emptyset \vdash_{\umode}\evalue{n_a}{\tau} }
    {(\varphi,\heap,\estar{\evalue{n}{\tptr{\tau}{\tmode}}}) \longrightarrow (\varphi,\heap,\evalue{n_a}{\tau})}

    \inferrule[S-AssignArrT]{\heap(\umode,n)=\evalue{n_a}{\tau_a}\\ 0 \in [n_l,n_h)
               \\  \emptyset;\heap ; \emptyset \vdash_{\umode}\evalue{n_1}{\tau} }
      {(\varphi,\heap,\eassign{\evalue{n}{\tallarrayptr{n_l}{n_h}{\tau}{\tmode}}}{\evalue{n_1}{\tau_1}}) \longrightarrow (\varphi,(\heap,\umode)[n \mapsto \evalue{n_1}{\tau_a}],\evalue{n_1}{\tau})}

@@

@@

    \inferrule[S-RetEnd]{}{(\varphi,\heap,\ret{x}{\mu}{\mu'}) \longrightarrow (\varphi,\heap,\mu')}

        \inferrule[S-Let]{}{(\varphi,\heap,\elet{x}{\evalue{n}{\tau}}{e}) \longrightarrow (\varphi,\heap,\ret{x}{\evalue{n}{\tau}}{e})}

@@

    \inferrule[S-Unchecked]{}{(\varphi,\heap,\eunchecked{\overline{x}}{\mu}) \longrightarrow (\varphi,\heap,\mu)}

@@

    \inferrule[S-Checked]{}{(\varphi,\heap,\echecked{\overline{x}}{\mu}) \longrightarrow (\varphi,\heap,\mu)}

    \inferrule[S-FunT]{ \Xi(\umode,n) = \tau\;(\evalue{\overline{x}}{\overline{\tau}})\;(\tmode,e)
                  \\ \emptyset;\heap ; \emptyset \vdash_{\umode}\evalue{n}{\tptr{\tau}{\tmode}}}
        {(\varphi,\heap,\ecall{\evalue{n}{(\tptr{\tau}{\tmode})}}{{\evalue{\overline{n_a}}{\overline{\tau_a}}}}) \longrightarrow
   (\varphi,\heap, \mathtt{let}\;\overline{x}={\evalue{\overline{n}}{(\overline{\tau}[\overline{n} / \overline{x}])}}\;\mathtt{in}\;\ecast{\tau[\overline{n} / \overline{x}]}{e})}

@@

\begin{figure}[t]
{\small
{\captionsetup[lstlisting]{margin = 8 mm}
  \begin{lstlisting}[xleftmargin=8 mm]
nt_array_ptr<char> safe_strcat
   (nt_array_ptr<char> dst : count(n),
    nt_array_ptr<char> src : count(0), int n) {
  int x = strlen(dst);
  int y = strlen(src);
  nt_array_ptr<char> c : count(n) =
    dyn_bounds_cast
           <nt_array_ptr<char>>(dst,count(n));
    // sets c == dst with bound n (not x)
  if (x+y < n) {
    for (int i = 0; i < y; ++i)
      *(c+x+i) = *(src+i);
    *(c+x+y) = '\0';
    return dst;
  }
  return null;
}
  \end{lstlisting}
}
}
\caption{Implementation of safe \code{strcat}}
\label{fig:strcat-ex}
\end{figure}

@@

\myparagraph{Checked and Tainted Pointer Operations}
%
The rules for pointer related operations---\textsc{S-DefC},
\textsc{S-DefT}, \textsc{S-AssignArrC}, \textsc{S-AssignArrT},
\textsc{S-DefNull}, and \textsc{S-Cast}.
The first five are deference and assignment operations---illustrate how the semantics checks bounds.
Rule \textsc{S-DefNull} transitions attempted null-pointer
dereferences to $\enull$, whereas \textsc{S-DefC} dereferences a \cmode-mode
non-null (single) pointer.
When $\enull$ is returned by the
computation relation, the evaluation relation halts the entire
evaluation with $\enull$ (using a rule not shown in Fig.~\ref{fig:c-context}); it
does likewise when $\ebounds$ is returned (see \Cref{sec:rem-semantics}).
\textsc{S-AssignArrC} assigns to an array as long as 0 (the point of
dereference) is within the bounds designated by the pointer's annotation
and strictly less than the upper bound.

\textsc{S-DefT} and \textsc{S-AssignArrT} are similar rules to \textsc{S-DefC} and \textsc{S-AssignArrC} for tainted pointers.
Any dynamic heap use of a tainted pointer requires a verification.
Performing such a verification equates to performing a type check for a pointer constant in \Cref{fig:const-type}.
We explain this shortly in \Cref{sec:type-system}.
For now, the verification step, e.g. $\umode;\emptyset;\heap ; \emptyset \vdash\evalue{n_a}{\tau}$ in \textsc{S-DefC},
means we verify that the value $n_a$ has type $\tau$ in the $\umode$ region heap.

Static casts of a literal $n\!:\!\tau'$ to a type $\tau$ are handled
by \textsc{S-Cast}. In a type-correct program, such casts are
confirmed safe by the type system no matter
if the target is a $\tmode$ or $\cmode$ pointer. To evaluate a cast, the rule
updates the type annotation on $n$. Before doing so, it must
``evaluate'' any variables that occur in $\tau$ according to their
bindings in $\varphi$. For example, if $\tau$ was
$\tarrayptr{0}{x+3}{\tint}{\cmode}$, then $\varphi(\tau)$ would
produce $\tarrayptr{0}{5}{\tint}{\cmode}$ if $\varphi(x) = 2$.
The full formalism, including \kw{struct}
and null-terminated bound widening pointer operations, is given in \Cref{app:main}.

@@

\myparagraph{Unchecked and Checked Blocks}
%
Semantically, $\euncheckedtext$ and $\echeckedtext$ blocks 
(\textsc{S-Unchecked} and \textsc{S-Checked}) act as a classical C block structures.
The interesting part is their type rules in \Cref{sec:type-system}.

\myparagraph{Binding and Function Calls}
%
The semantics handles variable scopes using the special $\erettext$
form. \textsc{S-Let} evaluates to a configuration whose stack
is $\varphi$ extended with a binding for $x$, and whose expression is
$\ret{x}{\evalue{n}{\tau}}{e})$ which keeps $\varphi$ unchanged, but
remember $x$ with the new value $\evalue{n}{\tau}$ in the scope $e$.
Every time when Evaluation proceeds on $e$ (Rule \textsc{S-RetCon}),
we install the current stack value $\evalue{n}{\tau}$ for $x$ in $\varphi$,
after one-step evaluation is completed, we restore the result $\varphi'(x)$ 
in the result $\ret{x}{\varphi'(x)}{e'})$.
This procedure continues until it becomes a literal
$n\!:\!\tau$, in which case \textsc{S-RetEnd} remove the $\kw{ret}$ frame and evaluates to
$n\!:\!\tau$.

Function calls are handled by \textsc{S-FunC} and \textsc{S-FunT},
for \cmode and \tmode mode function pointers, respectively. 
A call to function $f$ causes $f$'s
definition to be retrieved from $\Xi$,
% LEO: This is the first occurence of \Xi. It is explained,
% but reads a bit awkward as it's unclear this is its "introduction"
which maps function names to
function data $\tau\;(\evalue{\overline{x}}{\overline{\tau}})\;(\xi,e)$, where
$\tau$ is the return type, $(\evalue{\overline{x}}{\overline{\tau}})$
is the parameter list of variables and their types, 
$\xi$ determines the mode of the function, and $e$ is the
function body. 
Similar to \heap, the global function store $\Xi$ is also partitioned into
two parts ($\cmode$ and $\umode$ regions), each of which
maps addresses (integer literals) to the function data described above.
